### 2024-01-01 pnrs.tex
trying to advance

@@

\paragraph*{Context.} There has been recent interest in $\Rel$ and
$\Nat$-rational series exhibiting \emph{polynomial growth}
\cite{doueneau2021pebble,bojanczyk2022transducers}. This computational model,
in the counting case, can be traced back to a model introduced by
\textcite{schutzenberger1965finite}. It has been recently proven that one can
decide the growth rate of such functions, and decide whether they are aperiodic
in the case of $\Rel$-rational series \cite{LOPEZ23b}. It was conjectured in
the PhD thesis of Gaëtan Douéneau-Tabot, that the membership problem of
$\Nat$-rational series among $\Rel$-rational series was decidable under the
assumption of polynomial growth. This conjecture was based on the previously
introduced result of \textcite{KARH77} (\cref{karh:thm}), which we will
disprove in \cref{sec:c-example}. We then provide in \cref{sec:proof} a
corrected version of the theorem, together with a valid proof. As a
consequence, it does not invalidate the stated conjecture.

@@

\section{Deciding commutative $\Nat$-polyregular functions}
\label{sec:deciding}

@@

@@

>>>>

There has been recent interest in $\Rel$ and $\Nat$-rational series exhibiting
\emph{polynomial growth} \cite{doueneau2021pebble,bojanczyk2022transducers}.
This computational model, in the counting case, can be traced back to a model
introduced by \textcite{schutzenberger1965finite}. It has been recently proven
that one can decide the growth rate of such functions, and decide whether they
are aperiodic in the case of $\Rel$-rational series \cite{LOPEZ23b}. Isolating
a notion of aperiocidity for functions, and deciding this property is often a
non-trivial process \cite{droste2019aperiodic}. It was conjectured in the PhD
thesis of Gaëtan Douéneau-Tabot, that the membership problem of $\Nat$-rational
series among $\Rel$-rational series was decidable under the assumption of
polynomial growth. The semantic characterization would imply that
$\Nat$-rational series of polynomial growth that are aperiodic are precisely
$\Nat$-rational series of polynomial growth, that can be realised by
$\Rel$-rational series that are aperiodic. This conjecture was based on the
previously introduced result of \textcite[Theorem 3.3]{KARH77}, which we will
disprove in \cref{sec:c-example}. 
We then provide a corrected version of the theorem, and prove the conjecture
of Douéneau in the case of \emph{commutative} $\Nat$-rational series of
polynomial growth.

\section{Preliminaries}

@@

We then provide in \cref{sec:proof} a
corrected version of the theorem, together with a valid proof. As a
consequence, it does not invalidate the stated conjecture.

@@

\section{Effective procedures}
\label{sec:deciding}

@@

\subsection{Star-free $\Nat$-polyregular functions}

\begin{definition}[Ultimately polynomial]
\end{definition}

\begin{fact}
    Any $\Nat$-rational polynomial is a star-free polyregular function.
\end{fact}

\begin{fact}
    One can do anything on small values.
\end{fact}

Let us now conclude with a nice characterisation.
\begin{theorem}
    If
    $f$ is star-free $\Rel$-polyregular
    and $f$ is commutative $\Nat$-polyregular,
    then $f$ is star-free $\Nat$-polyregular.
\end{theorem}
\begin{proof}
    Idea. 
    Remark that $f$ is ultimately polynomial, and because $f$ 
    is commutative, this proves that $f$, after some rank,
    computes a polynomial. But this proves that $f$ is star-free
    $\Nat$-polyregular.
    (maybe we need to do an induction on the degree).
\end{proof}

\begin{corollary}
    One can decide if a commutative $\Nat$-polyregular function 
    is star-free.
\end{corollary}

@@

\begin{center}
    \begin{tabular}{c|cc}
        \toprule
        & \textbf{Commutative} & \textbf{General} \\
        \midrule
        \textbf{Commutative}
        and \textbf{Invertible} & commutative $\ZPoly$ & $\ZPoly$ \\
        \textbf{Commutative} & commutative $\NPoly$ & $\NPoly$ \\
        \textbf{General} & ? & polyregular 
    \end{tabular}
\end{center}
